Single-cell virus sequencing of influenza variants that trigger innate immunity

### 2018-09-15 paper/paper.tex
Split experiment figure into workflow and transcriptomics

\subsection{A system to identify and enrich for IFN+ positive cells}
A challenge in studying the induction of IFN by influenza is that infection with this virus only rarely triggers IFN expression at the single-cell level~\citep{killip2017single}.
For instance, we previously performed single-cell mRNA sequencing of hundreds of influenza-infected A549 cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}.
Only one of these cells expressed detectable type I or type III IFN (\FIG{IFNrare}A).
Importantly, the paucity of IFN+ cells identified in our prior single-cell mRNA sequencing was not due simply to a poor detection threshold, since $>$100 IFN transcripts were detected in the one IFN+ cell~\citep{russell2018extreme}.

The rareness of IFN induction by influenza infection is not merely a feature of A549 cells in culture.
Prior studies using fluorescent reporters have found that IFN expression is very rare at the single-cell level in influenza-infected mice at 24 hours post-infection~\citep{kallfass2013visualizing}.
In further support of these findings, we re-analyzed the single-cell mRNA sequencing of influenza-infected mice by \citet{steuerman2018dissection}.
As shown in \FIGSUPP[IFNrare]{mice}, many infected cells in the mice expressed interferon-stimulated genes, which are induced autocrine and paracrine signaling~\citep{stetson2006type,honda2006type}.
But only a handful of infected cells detectably expressed IFN itself, which is only induced by direct cellular detection of viral infection~\citep{stetson2006type,honda2006type}.

%%% start IFNrare figure %%%
\begin{figure}
\centerline{
{\bf \Large A}\includegraphics[width=0.25\textwidth,valign=t]{figures/IFN_stochastic/RussellVenn/venn_diagram.pdf}
\hspace{0.02\textwidth}
{\bf \Large B}\includegraphics[width=0.32\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
\hspace{0.02\textwidth}
{\bf \Large C} \includegraphics[width=0.18\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
}
\caption{
IFN expression is rare in influenza-infected cells.
{\bf (A)} Number of influenza-infected and IFN+ cells in our prior single-cell study of influenza infection of A549 cells~\citep{russell2018extreme}.
\FIGSUPP[IFNrare]{mice} shows that IFN+ cells were also rare in the single-cell mRNA-sequencing of influenza-infected mice by \citet{steuerman2018dissection}.
{\bf (B)} An A549 reporter cell line to identify single cells expressing IFN.
The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
We created reporter cell lines with type I (IFN-$\beta$) and type III (IFN-$\lambda$) promoters (\FIGDATA[IFNrare]{reporter_sequences}).
The reporters were efficiently activated when infected with an IFN-inducing strain of Sendai virus (\FIGSUPP[IFNrare]{reporter_validation}).
Activation of the type I and type III IFN reporters is highly correlated in single cells upon infection with influenza virus (\FIGSUPP[IFNrare]{type_I_vs_III}).
{\bf (C)}
Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the reporter cell line (see \FIGSUPP[IFNrare]{flow} for details).
The plot also shows uninfected cells, and cells infected with saturating amounts of Sendai virus.
The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
}
\label{fig:IFNrare}

\figsupp[Few cells express detectable IFN in the single-cell mRNA-sequencing of influenza-infected mice by \citet{steuerman2018dissection}.]
{A re-analysis of the data from \citet{steuerman2018dissection}'s single-cell mRNA-sequencing of cells from influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.
Specifically, we downloaded the data from \citet{steuerman2018dissection} and identified influenza-infected cells using calling criteria similar to those described in \citet{steuerman2018dissection}.
Here we show statistics for the cells from the influenza-infected wildtype C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
As shown in the left-most panel, the sequencing depth of \citet{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in \citet{russell2018extreme} and the current study, respectively).
An important caveat is that this low sequencing depth could lead to a simple failure to detect IFN transcripts in some cells.
Nonetheless, there is detectable expression of key interferon-stimulated gene (ISG) mRNAs in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, and only at low levels (one mRNA transcript detected).
The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
}
{\includegraphics[width=\textwidth]{figures/IFN_stochastic/SteuermanReanalysis/plots/p_mRNA_counts.pdf}}
\label{figsupp:mice}

@@

In order to efficiently identify and enrich cells that express IFN, we integrated IFN reporters into A549 cells (\FIG{IFNrare}B).
These reporters consisted of a type I (IFNB1) or type III (IFNL1) promoter driving expression of a cell-surface protein~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein.
Cells that activate IFN will express the cell-surface protein, enabling them to be enriched by magnetic-activated cell sorting (MACS).
The frequency of IFN+ cells can also be determined by flow cytometry for the fluorescent protein.
We validated that these A549 reporter cell lines were efficiently activated by infection with a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\FIGSUPP[IFNrare]{reporter_validation}).
We also used the reporters to show that expression of type I and type III IFN is highly correlated in single influenza-infected A549 cells (\FIGSUPP[IFNrare]{type_I_vs_III}), a result that is further validated by the single-cell mRNA sequencing described below.
Therefore, for the rest of this paper we will use the term ``IFN expression'' to refer to the combined expression of type I and III IFNs.

We then used one of the reporter cell lines to quantify the frequency of IFN induction in single A549 cells infected with the A/WSN/1933 (H1N1) strain of influenza.
As shown in \FIG{IFNrare}C, $\sim$0.5\% of infected cells activated the IFN reporter at 13 hours post-infection.
This frequency of IFN induction was roughly comparable to what we had observed in our prior single-cell mRNA sequencing of influenza-infected cells (\FIG{IFNrare}B), with the slightly higher rate in \FIG{IFNrare}C probably due to using a different viral stock and slightly later timepoint.

\subsection{new section}
All past single-cell transcriptomics of viral infected cells~\citep{russell2018extreme,zanini2018single,steuerman2018dissection,zanini2018virus} have focused on counting the number of viral transcripts rather than determining their full sequences.
We came up with a scheme to determine full genotypes of infecting viruses in IFN+ cells (\FIG{experiment}).

%%% start approach figure
\begin{figure}
\begin{fullwidth}

\hspace{0.04in} {\bf A} \includegraphics[width=0.95\linewidth, valign=t]{figures/WorkflowSchematic/SchematicForPaper.pdf}
\vspace{0.1in}

@@

\caption{
Single-cell sequencing of influenza infections that induce IFN.
{\bf (A)} Experimental workflow.
(1) IFN reporter A549 cells are infected with a mix of wildtype and synonymously barcoded viruses.
IFN+ cells are enriched by MACS (\FIGSUPP[experiment]{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
(2) The mRNAs from individual cells are tagged with cell-specific barcodes using a 10X Chromium device.
(3) The entire cellular transcriptomes are quantified using the standard single-cell mRNA-seq protocol, and viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio.
(4) The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
{\bf (B)} Number of cells for which transcriptomes were obtained.
From the transcriptomes with both human and canine transcripts, we estimate~\citep{bloom2018estimating} that $\approx$11\% of the captured cells are actually multiplets in which more than one cell was in an emulsion. 
{\bf (C)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
Green lines show the distribution of cellular mRNAs per cell, and the blue rug plot at the left of each panel shows the distribution of viral mRNAs per cell.
Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (possibly low-quality emulsions or multiplets), and are excluded from subsequent analyses.
{\bf (D)} Distribution across cells of the fraction of all mRNA derived from influenza.
Cells called as infected are in blue, while other cells are in green.
The inset shows the amount of viral mRNA among the human cells called as infected.
{\bf (E)} The number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
\FIGSUPP[experiment]{frac_has_gene} shows the frequency that each viral gene is detected in infected cells.
{\bf (F)} Relative expression of viral genes among infected cells.
{\bf (G)} The number of cells infected with wildtype virus, synonymously barcoded virus, or both.
From the cells infected with both viral barcodes, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected.
{\bf (H)} Fraction of cellular mRNA derived from IFN across all cells, faceted by whether the cells are infected.
Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line are classified as IFN+.
Many cells that do not express IFN itself still express ISGs (\FIGSUPP[experiment]{ISG}).
}
\label{fig:experiment}

\figsupp[MACS enrichment of IFN+ cells]
{
Example MACS enrichments of IFN+ influenza-infected cells.
The A549 cells with the IFN-$\beta$ LNGFR$\Delta$C-mNeonGreen reporter were infected with wildtype A/WSN/1933 influenza (two different viral stocks) at a target MOI of 0.1 TCID50 per cell.
After infection had proceeded for 12 hours, the cells were twice magnetically sorted for LNGFR$\Delta$C expression over magnetic columns as detailed in the methods for the single-cell sequencing experiment.
{\bf (A)}
After sorting, the populations were analyzed by flow cytometry for IFN expression using the mNeonGreen fluorescent protein.
The plots show the distribution of fluorescence in the original population, the flow-through from the first column, and the MACS-sorted positive population after two columns.
As indicated by the percentages shown for the original and MACS-sorted population, this process led to between 50- and 350-fold enrichment for IFN+ cells.
We expect that the IFN sorting for the actual single-cell sequencing led to similar enrichment, although we could not directly quantify this as the sorted cells in that case were immediately used for the sequencing and so could not be analyzed by flow cytometry.
{\bf (B)}
Analysis of expression of IFNB1 (relative to the housekeeping gene L32) by qPCR in the positive (IFN enriched) and negative (IFN depleted) populations from panel (A).
The qPCR validates a roughly 50- to 100-fold enrichment in total IFNB1 expression.
The qPCR was performed in quadruplicate (hence the four points for each sample).
}
{\includegraphics[width=\textwidth]{figures/MACS/MACS.pdf}}
\label{figsupp:MACS}

@@

\figsupp[Expression of interferon-stimulated genes (ISGs).]
{
For each cell, we quantified ISG expression as the total fraction of cellular mRNAs derived from four prototypical ISGs that are highly expressed A549 cells (IFIT1, ISG15, CCL5, and Mx1). 
{\bf (A)} The histograms show the distribution of ISG expression taken across infected (top) and uninfected (bottom) cells.
We heuristically classify as ISG+ cells with $>10^{-3}$ of their cellular mRNA from ISGs, and color these cells red.
Comparison to \FIG{experiment}H shows that substantially more cells are ISG+ than IFN+, both among infected and uninfected cells.
This is probably because paracrine signaling leads to ISG expression in some cells that are not themselves expressing IFN.
{\bf (B)} The correlation between the fraction of cellular mRNA derived from IFN and from ISGs.
Each point represents a single cell, and the Pearson correlation coefficient is shown.
IFN and ISG expression is more correlated for infected than uninfected cells, probably because in the latter the ISG expression is often due to paracrine signaling that does not induce expression of IFN itself.
But among both the infected and uninfected populations, there are many cells with high expression of ISGs and little expression of IFN, but very few cells that express high levels of IFN without also substantially expressing ISGs.
}
{\includegraphics[width=0.8\textwidth]{figures/single_cell_figures/p_isg.pdf}}
\label{figsupp:ISG}

@@

\end{fullwidth}
\end{figure}
%%% end approach figure

@@

%%% start genotypes figure %%%
\begin{figure}
\begin{fullwidth}
{\centering
\includegraphics[height=0.76\textheight]{figures/single_cell_figures/p_genotypes.pdf}
}
\caption{
Viral genotypes and infection outcomes in single cells.
Each row shows a single infected cell.
Arrows indicate the presence of a viral gene from a single viral barcode variant (light blue) or both barcode variants (dark blue).
Circles and boxes indicate mutations or indels as described in the legend.
Circle areas and box heights are proportional to the fraction of PacBio CCSs that report the mutation (see \FIGDATA[genotypes]{genotypes} for numerical data).
For dual-barcode infections, mutations / indels for the wildtype viral variant are on the top half of the arrow and those for the synonymously barcoded variant are on the bottom half. 
Green boxes to the left show the percent of all mRNA in that cell derived from virus.
Orange boxes show the percent of cellular mRNA derived from IFN, with boxes framed in orange indicating cells classified as IFN+ in \FIG{experiment}H.
}
\label{fig:genotypes}

@@

\figsupp[Number of PacBio circular consensus sequences and PCR strand exchange rate.]
{
The number of PacBio CCSs that passed quality-control steps and aligned to an influenza virus gene.
Note that these sequences were obtained using several PacBio runs, some of which were loaded with different amounts of the various viral genes in order to increase coverage on genes that were needed in order to obtain the full sequences of virions infecting cells.
Therefore, unlike the 10X transcript count data in \FIG{experiment}H, the numbers of CCSs for different genes should \emph{not} be taken as an indicator of their abundance in the infected cells.
Especially for the polymerase genes (PB2, PB1, and PA), many of the CCSs corresponded to genes with internal deletions, since these shorter forms of the genes were presumably preferentially amplified during PCR.
Therefore, the plot is faceted by the number of CCSs for any length of the gene, and for full-length genes.
Note that the disproportionate sequencing of the shorter internally deleted genes should not greatly affect the genotype calling in \FIG{genotypes} since UMIs were used to collapse duplicate sequences for the same gene, and cell barcodes to collapse duplicate sequences from the same cell.
The bars are colored by whether the sequence can be identified as being derived from the wildtype viral variant, the synonymously barcoded variant, or represents a mixed barcode molecule (see \FIGSUPP[genotypes]{StrandExchange}).
From the frequencies of these different forms, we estimate~\citep{bloom2018estimating} that 5.7\% of molecules are chimeric due to PCR strand exchange.
Note that about half of these PCR chimeras could be identified by the presence of mixed viral barcodes and removed from subsequent analyses, leaving $\sim$3\% un-identified chimeras.
Note also that for some molecules (mostly polymerase genes with internal deletions) one of the barcode sites was deleted from the molecule and so the barcode identity could only be partially called.
A negligible number of molecules have low-accuracy sequence or unexpected nucleotide identities at the sites of the viral barcode.
}
{\includegraphics[width=\textwidth]{figures/pacbio_single_cell_figures/ccs_per_viralbarcode.pdf}}
\label{figsupp:CCSs}

>>>>

\subsection{A system to identify and enrich for rare IFN+ positive cells}
A challenge in studying the induction of IFN by influenza is that infection with this virus only rarely triggers IFN expression at the single-cell level~\citep{killip2017single}.
For instance, we previously performed single-cell transcriptomics on hundreds of influenza-infected A549 cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}.
Only 1 of 368 infected cells expressed detectable type I or type III IFN (\FIG{IFNrare}A).
Importantly, the paucity of IFN+ cells identified in our prior single-cell transcriptomics was not due simply to a poor detection threshold, since $>$100 IFN transcripts were detected in the one IFN+ cell~\citep{russell2018extreme}.

The rareness of IFN induction by influenza infection is not merely a feature of A549 cells in culture.
Prior studies using fluorescent reporters have found that IFN expression is very rare at the single-cell level in influenza-infected mice at 24 hours post-infection~\citep{kallfass2013visualizing}.
In further support of these findings, we re-analyzed the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection}.
As shown in \FIGSUPP[IFNrare]{mice}, many infected cells in the mice expressed IFN-stimulated genes, which are induced autocrine and paracrine signaling~\citep{stetson2006type,honda2006type}.
But only a handful of infected cells detectably expressed IFN itself, which is only induced by direct cellular detection of viral infection~\citep{stetson2006type,honda2006type}.

%%% start IFNrare figure %%%
\begin{figure}
\centerline{
{\bf \Large A}\includegraphics[width=0.25\textwidth,valign=t]{figures/IFN_stochastic/RussellVenn/venn_diagram.pdf}
\hspace{0.02\textwidth}
{\bf \Large B}\includegraphics[width=0.32\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
\hspace{0.02\textwidth}
{\bf \Large C} \includegraphics[width=0.18\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
}
\caption{
IFN expression is rare in influenza-infected cells.
{\bf (A)} Number of infected and IFN+ cells in our prior single-cell transcriptomics of influenza-infected A549 cells~\citep{russell2018extreme}.
\FIGSUPP[IFNrare]{mice} shows that IFN+ cells were also rare in the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection}.
{\bf (B)} An A549 reporter cell line to identify single cells expressing IFN.
The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
We created reporter cell lines with type I (IFN-$\beta$) and type III (IFN-$\lambda$) promoters (\FIGDATA[IFNrare]{reporter_sequences}).
The reporters were efficiently activated when infected with an IFN-inducing strain of Sendai virus (\FIGSUPP[IFNrare]{reporter_validation}).
Activation of the type I and type III IFN reporters is highly correlated in single cells upon infection with influenza virus (\FIGSUPP[IFNrare]{type_I_vs_III}).
{\bf (C)}
Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the reporter cell line (see \FIGSUPP[IFNrare]{flow} for details).
The plot also shows uninfected cells, and cells infected with saturating amounts of Sendai virus.
The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
}
\label{fig:IFNrare}

\figsupp[Few cells express detectable IFN in the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection}.]
{A re-analysis of the data from \citet{steuerman2018dissection}'s single-cell transcriptomics of cells from influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.
Specifically, we downloaded the data from \citet{steuerman2018dissection} and identified influenza-infected cells using calling criteria similar to those described in \citet{steuerman2018dissection}.
Here we show statistics for the cells from the influenza-infected wildtype C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
As shown in the left-most panel, the sequencing depth of \citet{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in \citet{russell2018extreme} and the current study, respectively).
An important caveat is that this low sequencing depth could lead to a simple failure to detect IFN transcripts in some cells.
Nonetheless, there is detectable expression of key interferon-stimulated gene (ISG) mRNAs in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, and only at low levels (one mRNA transcript detected).
The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
}
{\includegraphics[width=\textwidth]{figures/IFN_stochastic/SteuermanReanalysis/plots/p_mRNA_counts.pdf}}
\label{figsupp:mice}

@@

In order to efficiently identify and enrich cells that express IFN, we integrated IFN reporters into the genome of A549 cells (\FIG{IFNrare}B).
These reporters consisted of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein.
Cells that activate IFN will express the cell-surface protein, enabling them to be enriched by magnetic-activated cell sorting (MACS).
The frequency of IFN+ cells can also be determined by flow cytometry for the fluorescent protein.
We validated that the IFN reporters were efficiently activated by infection with a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\FIGSUPP[IFNrare]{reporter_validation}).
We also used the reporters to show that expression of type I and type III IFN is highly correlated in single influenza-infected A549 cells (\FIGSUPP[IFNrare]{type_I_vs_III}), a result that is further validated by our single-cell transcriptomics described below.
Therefore, for the rest of this paper we will use the term ``IFN expression'' to refer to the combined expression of type I and III IFNs.

We then used the reporter cells to quantify the frequency of IFN induction in single A549 cells infected with the A/WSN/1933 (H1N1) strain of influenza (hereafter referred to as ``WSN'').
As shown in \FIG{IFNrare}C, $\sim$0.5\% of infected cells activated the IFN reporter at 13 hours post-infection.
This frequency of IFN induction was roughly comparable to what we had observed in our prior single-cell transcriptomics of influenza-infected cells (\FIG{IFNrare}B), with the slightly higher rate in \FIG{IFNrare}C probably due to using a different viral stock and slightly later timepoint.

\subsection{Simultaneous transcriptomics and virus-sequencing of single infected cells}
We next set out to determine whether features of the infecting virions are responsible for the heterogeneous outcome of influenza infection in single cells, including the rare induction of IFN.
Standard single-cell transcriptomic approaches are inadequate to address this question, since they typically just use 3'-end sequencing of poly-adenylated mRNAs to determine the abundance of each transcript~\citep{klein2015droplet, macosko2015highly, zheng2017massively, cao2017comprehensive, gierahn2017seq}.
But simply measuring the abundance of viral transcripts is insufficient for our purposes: we want to know the full sequences of all viral genes in each infected cell, so that we can determine if features of these genes (e.g., mutations) contribute to heterogeneity in infection outcome.

%%% start workflow figure
\begin{figure}
\begin{fullwidth}

\includegraphics[width=\linewidth, valign=t]{figures/WorkflowSchematic/SchematicForPaper.pdf}

\caption{
Approach for simultaneous transcriptomics and viral sequencing of single influenza-infected cells that express IFN.
{\bf (A)}
IFN reporter A549 cells are infected with a mix of wildtype and synonymously barcoded viruses.
IFN+ cells are enriched by MACS (\FIGSUPP[workflow]{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
{\bf (B)}
The mRNAs from individual cells are tagged with cell-specific barcodes using a 10X Chromium device.
{\bf (C)}
The entire cellular transcriptomes are quantified using standard single-cell 3'-end Illumina sequencing, and 
{\bf (D)}
viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio.
{\bf (E)}
The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
}
\label{fig:workflow}

\figsupp[MACS enrichment of IFN+ cells.]
{
Example MACS enrichments of IFN+ influenza-infected cells.
The A549 cells with the IFN-$\beta$ LNGFR$\Delta$C-mNeonGreen reporter were infected with wildtype WSN influenza (two different viral stocks) at a target MOI of 0.1 TCID50 per cell.
After infection had proceeded for 12 hours, the cells were twice magnetically sorted for LNGFR$\Delta$C expression over magnetic columns as detailed in the methods for the single-cell sequencing experiment.
{\bf (A)}
After sorting, the populations were analyzed by flow cytometry for IFN expression using the mNeonGreen fluorescent protein.
The plots show the distribution of fluorescence in the original population, the flow-through from the first column, and the MACS-sorted positive population after two columns.
As indicated by the percentages shown for the original and MACS-sorted population, this process led to between 50- and 350-fold enrichment for IFN+ cells.
We expect that the IFN sorting for the actual single-cell sequencing led to similar enrichment, although we could not directly quantify this as the sorted cells in that case were immediately used for the sequencing and so could not be analyzed by flow cytometry.
{\bf (B)}
Analysis of expression of IFNB1 (relative to the housekeeping gene L32) by qPCR in the positive (IFN enriched) and negative (IFN depleted) populations from panel (A).
The qPCR validates a roughly 50- to 100-fold enrichment in total IFNB1 expression.
The qPCR was performed in quadruplicate (hence the four points for each sample).
}
{\includegraphics[width=\textwidth]{figures/MACS/MACS.pdf}}
\label{figsupp:MACS}

\end{fullwidth}
\end{figure}
%%% end workflow figure

We developed the experimental approach in \FIG{workflow} to simultaneously determine the entire transcriptome and the full sequences of the infecting virions in single cells.
First, we generated a stock of virus that consisted of a mix of two variants of the WSN strain: the wildtype virus and a ``synonymously barcoded'' variant that contains two engineered synonymous mutations near each end of each gene (\FIGDATA[workflow]{virus_seqs}).
We have previously shown that synonymously barcoding viral genes near the 3' end of the mRNA enables quantification of the co-infection rate in single-cell transcriptomic data, since about half the cells infected by multiple virions express a mix of wildtype and synonymously barcoded viral transcripts~\citep{russell2018extreme}. 
New to this study, we also added synonymous barcodes near the 5' end of the mRNA; as explained below, having barcodes at both ends provides an important control for PCR artifacts during full-length sequencing of viral transcripts.

We next used this viral stock to infect A549 cells carrying a type I IFN reporter (\FIG{transcriptomics}A), using a viral dose that we expected to lead to transcriptionally active viral infection in slightly less than half the cells.
From 12 to 13 hours post-infection, we used magnetic columns to enrich for cells that expressed the cell-surface protein driven by the IFN reporter.
Based on pilot experiments, we expected this MACS to enrich the rare IFN+ cells to about 5\% to 30\% of the total population (\FIGSUPP[workflow]{MACS}; the MACS enrichment efficiency is imperfect so many sorted cells are still IFN-).
In order to ensure the presence of IFN- cell, we also added back non-sorted cells to $\sim$10\% of the total.
We also added uninfected canine cells to $\sim$5\% of the total to serve as an internal control for multiplets and to estimate the background amount of viral mRNA observed in the transcriptomes of truly uninfected cells.

%%% start transcriptomics figure
\begin{figure}
\begin{fullwidth}

@@

\caption{
Single-cell transcriptomics of influenza-infected cells enriched for IFN positivity.
{\bf (A)} 
Number of cells for which transcriptomes were obtained.
From the transcriptomes with both human and canine transcripts, we estimate~\citep{bloom2018estimating} that $\approx$11\% of the captured cells are actually multiplets in which more than one cell was in an emulsion. 
{\bf (B)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
Green lines show the distribution of cellular mRNAs per cell, and the blue rug plot at the left of each panel shows the distribution of viral mRNAs per cell.
Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (possibly low-quality emulsions or multiplets), and are excluded from subsequent analyses.
{\bf (C)} Distribution across cells of the fraction of all mRNA derived from influenza.
Cells called as infected are in blue, while other cells are in green.
The inset shows the amount of viral mRNA among the human cells called as infected.
{\bf (D)} The number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
\FIGSUPP[transcriptomics]{frac_has_gene} shows the frequency that each viral gene is detected in infected cells.
{\bf (E)} Relative expression of viral genes among infected cells.
{\bf (F)} The number of cells infected with wildtype virus, synonymously barcoded virus, or both.
From the cells infected with both viral barcodes, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected.
{\bf (G)} Fraction of cellular mRNA derived from IFN across all cells, faceted by whether the cells are infected.
Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line are classified as IFN+.
Many cells that do not express IFN itself still express ISGs (\FIGSUPP[transcriptomics]{ISG}).
}
\label{fig:transcriptomics}

@@

\figsupp[Expression of interferon-stimulated genes (ISGs).]
{
For each cell, we quantified ISG expression as the total fraction of cellular mRNAs derived from four prototypical ISGs that are highly expressed A549 cells (IFIT1, ISG15, CCL5, and Mx1). 
{\bf (A)} The histograms show the distribution of ISG expression taken across infected (top) and uninfected (bottom) cells.
We heuristically classify as ISG+ cells with $>10^{-3}$ of their cellular mRNA from ISGs, and color these cells red.
Comparison to \FIG{transcriptomics}G shows that substantially more cells are ISG+ than IFN+, both among infected and uninfected cells.
This is probably because paracrine signaling leads to ISG expression in some cells that are not themselves expressing IFN.
{\bf (B)} The correlation between the fraction of cellular mRNA derived from IFN and from ISGs.
Each point represents a single cell, and the Pearson correlation coefficient is shown.
IFN and ISG expression is more correlated for infected than uninfected cells, probably because in the latter the ISG expression is often due to paracrine signaling that does not induce expression of IFN itself.
But among both the infected and uninfected populations, there are many cells with high expression of ISGs and little expression of IFN, but very few cells that express high levels of IFN without also substantially expressing ISGs.
}
{\includegraphics[width=0.8\textwidth]{figures/single_cell_figures/p_isg.pdf}}
\label{figsupp:ISG}

@@

\end{fullwidth}
\end{figure}
%%% end transcriptomics figure

@@

%%% start genotypes figure %%%
\begin{figure}
\begin{fullwidth}
{\centering
\includegraphics[height=0.76\textheight]{figures/single_cell_figures/p_genotypes.pdf}
}
\caption{
Viral genotypes and infection outcomes in single cells.
Each row shows a single infected cell.
Arrows indicate the presence of a viral gene from a single viral barcode variant (light blue) or both barcode variants (dark blue).
Circles and boxes indicate mutations or indels as described in the legend.
Circle areas and box heights are proportional to the fraction of PacBio CCSs that report the mutation (see \FIGDATA[genotypes]{genotypes} for numerical data).
For dual-barcode infections, mutations / indels for the wildtype viral variant are on the top half of the arrow and those for the synonymously barcoded variant are on the bottom half. 
Green boxes to the left show the percent of all mRNA in that cell derived from virus.
Orange boxes show the percent of cellular mRNA derived from IFN, with boxes framed in orange indicating cells classified as IFN+ in \FIG{transcriptomics}G.
}
\label{fig:genotypes}

@@

\figsupp[Number of PacBio circular consensus sequences and PCR strand exchange rate.]
{
The number of PacBio CCSs that passed quality-control steps and aligned to an influenza virus gene.
Note that these sequences were obtained using several PacBio runs, some of which were loaded with different amounts of the various viral genes in order to increase coverage on genes that were needed in order to obtain the full sequences of virions infecting cells.
Therefore, unlike the 10X transcript count data in \FIG{transcriptomics}, the numbers of CCSs for different genes should \emph{not} be taken as an indicator of their abundance in the infected cells.
Especially for the polymerase genes (PB2, PB1, and PA), many of the CCSs corresponded to genes with internal deletions, since these shorter forms of the genes were presumably preferentially amplified during PCR.
Therefore, the plot is faceted by the number of CCSs for any length of the gene, and for full-length genes.
Note that the disproportionate sequencing of the shorter internally deleted genes should not greatly affect the genotype calling in \FIG{genotypes} since UMIs were used to collapse duplicate sequences for the same gene, and cell barcodes to collapse duplicate sequences from the same cell.
The bars are colored by whether the sequence can be identified as being derived from the wildtype viral variant, the synonymously barcoded variant, or represents a mixed barcode molecule (see \FIGSUPP[genotypes]{StrandExchange}).
From the frequencies of these different forms, we estimate~\citep{bloom2018estimating} that 5.7\% of molecules are chimeric due to PCR strand exchange.
Note that about half of these PCR chimeras could be identified by the presence of mixed viral barcodes and removed from subsequent analyses, leaving $\sim$3\% un-identified chimeras.
Note also that for some molecules (mostly polymerase genes with internal deletions) one of the barcode sites was deleted from the molecule and so the barcode identity could only be partially called.
A negligible number of molecules have low-accuracy sequence or unexpected nucleotide identities at the sites of the viral barcode.
}
{\includegraphics[width=\textwidth]{figures/pacbio_single_cell_figures/ccs_per_viralbarcode.pdf}}
\label{figsupp:CCSs}

\figsupp[Number of cells for which genotype(s) of infecting viruses were completely determined.]
{Cells for which we could determine the full sequences of all genes expressed by the infecting virus(es).
{\bf (A)} We could call the complete genotypes of the infecting virus(es) for the majority of cells infected with just a single viral barcode variant, but only a minority of cells co-infected with both viral barcodes.
{\bf (B)} The cells for which we could call complete viral genotypes tended to have higher expression of viral mRNAs than cells for which we could not call complete genotypes.
This makes sense, as cells with more viral mRNA are more likely to have their viral genes captured in the PacBio sequencing, which is only able to capture a small fraction of the total transcripts identified by the 3'-sequencing of the 10X platform.
The lower calling rate for dual-barcode co-infections relative to single barcode infectionis is probably because these co-infections have more viral genes that must be sequenced (potentially a copy of each viral gene from each viral variant), increasing the chances that one of these genes is missed by the PacBio sequencing. 
An important implication of this plot is that the cells for which we call complete viral genotypes are \emph{not} a random subsampling of all infected cells in the experiment, but are rather enriched for cells that have high levels of viral mRNA and do not have dual-barcode viral infections.
Note also that this plot is limited to the cells that were called as infected (\FIG{transcriptomics}C) and could clearly be classified as IFN- or IFN+ (\FIG{transcriptomics}G).
}
{\includegraphics[width=0.8\textwidth]{figures/single_cell_figures/p_cells_complete.pdf}}
\label{figsupp:ncells}
